---
title: "Demographic and Employment Trends Among Elderly U.S. Immigrant Populations"
output: html_notebook
---

## Library packages

```{r}
# ipsumr packages for reading ddi and modifying the dataset
# dplyr for data manipulation 
library(ggplot2)
library(ipumsr)
library(dplyr)
library(haven) 
```

# Loading Data

```{r}
# Example data
# reading data
ddi <- read_ipums_ddi("highered_00001.xml")
data <- read_ipums_micro(ddi)
```

```{r}
#Variables within the data set
ddi$var_info
```

# BTHUS indicates wether an observation was born in US or not

```{r}
ipums_val_labels(ddi, BTHUS)
# 0 is observations not born in United States
# 1 is observations in born in United States
# 99 is observations is neither not defined or unnamed
```

# LFSAT is the Labor Force Status

```{r}
ipums_val_labels(ddi, LFSTAT)
#1 is indicating employed
#2 is indicating unepmloyed
#3 is not in the labor force
```

# 

```{r}
ipums_var_desc(ddi, AGE)
ipums_var_desc(ddi, GENDER)
ipums_val_labels(ddi, GENDER)
ipums_val_labels(ddi, RACETH)
ipums_var_desc(ddi, DGRDG)
ipums_val_labels(ddi, DGRDG)
```

# Analyzing elder immigrrants(60 and above)

```{r}
# Filter for elder immigrants (age >= 60 and not born in the U.S.)
elder_immigrants1 <- data %>%
  filter(AGE >= 60, BTHUS != 1)  # Adjust "United States" based on how it's encoded

```

```{r}
head(elder_immigrants1)
```

```{r}
# Summarize labor force status
work_status_summary <- elder_immigrants1 %>%
  group_by(LFSTAT) %>%
  summarize(count = n())  # Count number of individuals in each labor force status category

# View the summary
print(work_status_summary)

#LFSTAT
# 1/ Employed
# 2/ Umeployed
# 3/ Not in the labor force

```

```{r}
#Summarize labor work force status for the elder_immigrants2

work_status_summary2 <- elder_immigrants2 %>% group_by(LFSTAT) %>%
summarize(count = n())


print(work_status_summary2)

#1 is Employed
#2 is Umeployed
#3 is Not in Work Force
```

```{r}
#duplicate
#Summarize labor work force status for the elder_immigrants2

work_status_summary <- elder_immigrants1 %>% group_by(LFSTAT) %>%
summarize(count = n())


print(work_status_summary)

#1 is Employed
#2 is Umemployed
#3 is Not in Work Force
```

```{r}
# Convert LFSTAT to a factor
elder_immigrants2 <- elder_immigrants2 %>%
  mutate(LFSTAT = as_factor(LFSTAT))

# Check the unique values of LFSTAT after conversion
unique(elder_immigrants$LFSTAT)
```

```{r}
# Summarize labor force status using the original numeric values
work_status_summary2 <- elder_immigrants2 %>%
  group_by(LFSTAT) %>%
  summarize(count = n(), .groups = 'drop')

# View the summary
print(work_status_summary)

```

```{r}
# Bar plot of labor force status distribution using numeric values
ggplot(work_status_summary, aes(x = factor(LFSTAT, levels = c(1, 2, 3)), y = count, fill = factor(LFSTAT))) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = c("1" = "Employed", "2" = "Unemployed", "3" = "Not in the labor force")) +
  labs(title = "Labor Force Status of Elder Immigrants", x = "Labor Force Status", y = "Count") +
  theme_minimal()

```

```{r}

```

### Elderly Workforce by Gender

```{r}
# Summarize labor force status by gender
work_status_by_gender <- elder_immigrants1 %>%
  group_by(GENDER, LFSTAT) %>%
  summarize(count = n())

#print(work_status_by_gender)


# Recode GENDER directly in the ggplot call
ggplot(work_status_by_gender, aes(x = factor(LFSTAT, levels = c("1", "2", "3")), 
                                  y = count, 
                                  fill = factor(GENDER, labels = c("Female", "Male")))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_discrete(labels = c("1" = "Employed", "2" = "Unemployed", "3" = "Not in the labor force")) +
  labs(title = "Labor Force Status of Elder Immigrants by Gender", 
       x = "Labor Force Status", 
       y = "Count", 
       fill = "Gender") +
  theme_minimal()
```

# As the graph shows, male elder immigrants are four times more likely to be in the workforce than female elder immigrants.

### Analyzing race/ethnicity impact on the labor force

```{r}
# Summarize labor force status by race/ethnicity
work_status_by_race <- elder_immigrants1 %>%
  group_by(RACETH, LFSTAT) %>%
  summarize(count = n(), .groups = 'drop')
```

```{r}
# Check unique values for RACETH 
unique(work_status_by_race$RACETH)


```

```{r}
# Recode RACETH as a factor with specified labels
work_status_by_race <- work_status_by_race %>%
  mutate(RACETH = factor(RACETH, levels = c(1, 2, 3, 4),
                         labels = c("Asian", "White", "Under-represented minorities", "Other")))

# Plot with the RACETH
ggplot(work_status_by_race, aes(x = factor(LFSTAT, levels = c("1", "2", "3")), 
                                y = count, 
                                fill = RACETH)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_discrete(labels = c("1" = "Employed", "2" = "Unemployed", "3" = "Not in the labor force")) +
  labs(title = "Labor Force Status of Elder Immigrants by Race", 
       x = "Labor Force Status", 
       y = "Count", 
       fill = "Race/Ethnicity") +
  theme_minimal()

```

# Asian elder immigrants may have higher labor force participation compared to under-represented minorities in your dataset.

# \* The racial/ethnic variables in the dataset lacked more definitive categories, making it impossible to differentiate between Hispanic, Black, or mixed groups.

### #Educational Attainment impact on labor forrce

```{r}
elder_immigrants1 <- elder_immigrants1 %>%
  mutate(DGRDG = recode(DGRDG, 
                        `1` = "Bachelor's", 
                        `2` = "Master's", 
                        `3` = "Doctorate", 
                        `4` = "Professional"))
```

```{r}
# Summarize labor force status by education level
work_status_by_education <- elder_immigrants1 %>%
  group_by(DGRDG, LFSTAT) %>%
  summarize(count = n())
```

```{r}
head(work_status_by_education)

ipums_var_desc(ddi, DGRDG)
ipums_val_labels(ddi, DGRDG)

#1 = Bachelors
#2 = Masters
#3 = Doctorate
#4 = Professional 

```

```{r}
ggplot(work_status_by_education, 
       aes(x = recode(as_factor(DGRDG), 
                      "1" = "Bachelor's", 
                      "2" = "Master's", 
                      "3" = "Doctorate", 
                      "4" = "Professional"), 
           y = count, 
           fill = recode(as_factor(LFSTAT), 
                         "1" = "Employed", 
                         "2" = "Unemployed", 
                         "3" = "Not in the Labor Force"))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Labor Force Status of Elder Immigrants by Education Level", 
       x = "Highest Degree", 
       y = "Count", 
       fill = "Labor Force Status") +  # Add a legend title for clarity
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

# Elder immigrants worker who have Doctorates seems to be the biggest group in the data

```{r}
# Summarize labor force status by age groups
work_status_by_age <- elder_immigrants1 %>%
  group_by(cut(AGE, breaks = seq(65, 100, by = 5)), LFSTAT) %>%
  summarize(count = n())
```

```{r}
head(work_status_by_age)
```

## Overview on immigrants workers from age of 30 to 50

```{r}
#Making a new data set for observation from 50 and under, and not born in the US.

demo <- data %>%
  filter(AGE <= 50,BTHUS != 1 )
```

```{r}
# Summarize labor force status by age group, now including 30 to 50
work_status_by_age_30_to_50 <- demo %>%
  filter(AGE >= 30 ) %>%  # Filtering for ages 30 and above
  mutate(age_group = cut(AGE, breaks = seq(30, 50, by = 5))) %>%
  group_by(age_group, LFSTAT) %>%
  summarize(count = n())

# View the summary to verify counts by age group and labor force status
print(work_status_by_age_30_to_50)

```

```{r}
ggplot(work_status_by_age_30_to_50, aes(x = age_group, 
                               y = count, 
                               fill = factor(LFSTAT, 
                                             levels = c("1", "2", "3"), 
                                             labels = c("Employed", "Unemployed", "Not in Labor Force")))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Labor Force Status by Age Group (50 and Below) for Immigrrants", 
       x = "Age Group", 
       y = "Count", 
       fill = "Labor Force Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Majority of immigrant from age 50 and below are employed.
